---
title: "Fish"
author: "Thomas Lumley"
date: "2023-04-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setting up the survival times

```{r cars}
fish<-read.csv("../../data/release_mortality.csv")
```

Most of the `Control` fish are missing `capture_time` and `total_transfer_time`

```{r}
with(fish, table(capture_time=="", treatment))
with(fish, table(total_transfer_time=="", treatment))
with(fish, table(wait_time=="", treatment))
```

so we really have to start with `net_release_time`, which is only missing for 6 fish, or `pickup_datetime`

```{r}
with(fish, table(net_release_time=="", treatment))
with(fish, table(pickup_datetime=="", treatment))
```

Transform these to proper date/time objects

```{r}
fish$pickup<-as.POSIXct(fish$pickup_datetime, format="%Y-%m-%d %H:%M")
fish$retrieval<-as.POSIXct(paste(fish$retrieval_date,fish$retrieval_time), format="%Y-%m-%d %H:%M")
fish$release_date<-substr(fish$pickup_datetime,1,10)
fish$release<-as.POSIXct(paste(fish$release_date,fish$net_release_time), format="%Y-%m-%d %H:%M")
library(survival)
mean(fish$release-fish$pickup,na.rm=TRUE)
max(fish$release-fish$pickup,na.rm=TRUE)

plot(survfit(Surv(retrieval-release, fish_retrieval_fate=="Died")~1,data=fish),mark.time=TRUE,xscale=60,xlab="Hours since release",ylab="Proportion alive")
plot(survfit(Surv(retrieval-release, fish_retrieval_fate=="Died")~treatment,data=fish),col=c("orange","blue"),xscale=60,xlab="Hours since release",ylab="Proportion alive")
with(fish, table(treatment, fish_retrieval_fate))
```

So, all the `Control` fish lived, 83% of the `Recreational` fish lived, and the deaths happened over a short period of time before any of the censoring, so that survival analysis doesn't give much benefit over just a binary analysis.

## Comparisons

### Hook location

Since we know all `Control` fish were lip-hooked and none died, we can either exclude them completely or put them in the `Lip` category. The survival rate for `Control` fish is clearly better than for lip-hooked `Control` fish so it makes sense to exclude the controls.

```{r}
table(fish$capture_hook_site)
with(fish, table(capture_hook_site, fish_retrieval_fate))
recr_fish<-subset(fish, treatment=="Recreational")
recr_fish$hook_site<-factor(recr_fish$capture_hook_site, c("Lip","Gut","Foul"))
coxph(Surv(retrieval-release, fish_retrieval_fate=="Died")~hook_site,data=recr_fish)
plot(survfit(Surv(retrieval-release, fish_retrieval_fate=="Died")~hook_site,data=recr_fish),col=c("sienna","red","black"), xscale=60,xlab="Hours since release",ylab="Proportion alive")
legend("bottomleft",lty=1,col=c("sienna","red","black"),legend=levels(recr_fish$hook_site),bty="n")

summary(glm(I(fish_retrieval_fate=="Died")~hook_site, data=recr_fish,family=binomial(log)))
```

There is strong evidence that gut-hooked and foul-hooked fish are more likely to die than lip-hooked fish. The rate of death is about 8 times as high for gut-hooked fish and about twice as high for foul-hooked fish. Alternatively, modelling relative risks of death and ignoring time, gut-hooked fish are about `exp(1.6)` or 5 times as likely to die and foul-hooked fish about `exp(0.68)` or twice as likely to die.

### Capture depth

```{r}
library(beeswarm)
beeswarm(capture_depth_m~treatment,data=fish,col=1:2,cex=0.2)
beeswarm(capture_depth_m~capture_hook_site,data=fish,col=1:4,cex=0.2)
beeswarm(capture_depth_m~capture_depth_range,data=recr_fish,col=1:4,cex=0.2)
```

The `Control` fish are very different in capture depth distribution (and we know they all lived), so again we can drop them from the analysis.

```{r}
coxph(Surv(retrieval-release, fish_retrieval_fate=="Died")~capture_depth_range,data=recr_fish)
plot(survfit(Surv(retrieval-release, fish_retrieval_fate=="Died")~capture_depth_range,data=recr_fish),col=c("navyblue","royalblue","lightblue"),lwd=2,xscale=60,xlab="Hours since release",ylab="Proportion alive")
legend("bottomleft",lty=1,col=c("navyblue","royalblue","lightblue"),legend=c("Deep","Medium","Shallow"),bty="n",lwd=2)

```

The rate of death at `Deep` and `Medium` ranges is similar (hazard ratio 0.9 for `Medium` vs `Deep`) but the rate of death at `Shallow` capture range is approximately 5 times lower. There is very strong evidence for this difference.

```{r}
summary(glm(I(fish_retrieval_fate=="Died")~capture_depth_range,data=recr_fish, family=binomial(log)))
```

The same conclusions are found with a binary analysis of fish fate: at most a small difference between `Deep` and `Medium` but nearly 5 times lower risk of death at `Shallow`

### Depth in more detail

Here we fit a linear term to depth within the `Shallow` range:

```{r}
coxph(Surv(retrieval-release, fish_retrieval_fate=="Died")~capture_depth_m, data=subset(recr_fish, capture_depth_range=="Shallow"))

```

There is strong evidence that within the `Shallow` range, mortality increases with depth

```{r}
library(splines)
m<-coxph(Surv(retrieval-release, fish_retrieval_fate=="Died")~ns(capture_depth_m,3), data=subset(recr_fish, capture_depth_range="Shallow"))
termplot(m, ylab="log hazard",se=TRUE, rug=TRUE)
```

It appears that the mortality rate increases until about 15-20m depth and then becomes roughly constant.

### Combining the factors

```{r}
m2<-coxph(Surv(retrieval-release, fish_retrieval_fate=="Died")~capture_depth_range*hook_site, data=recr_fish)
anova(m2)
summary(m2)
```

When the factors are combined there is strong evidence that both `capture_depth_range` and `hook_site` affect mortality, and also that the effect of `hook_site` is different at different depths. Specifically, there was much more excess mortality among gut-hooked fish at the `Shallow` depth: the hazard ratio was estimated as 4.75 for `Deep` captures (and approximately the same for `Medium`), but 4.75x14.7=70 times higher at `Shallow` depths.

```{r fig.height=6}
recr_fish$site_depth<-with(recr_fish, interaction(hook_site, ifelse(capture_depth_range=="Shallow","Shallow","not Shallow"))) 
plot(survfit(Surv(retrieval-release, fish_retrieval_fate=="Died")~site_depth,data=recr_fish),col=c("sienna","red","black"),lty=c(1,1,1,2,2,2),xscale=60,xlab="Hours since release",ylab="Proportion alive")
legend("bottomleft",col=c("sienna","red","black"),,lty=c(1,1,1,2,2,2),legend=levels(recr_fish$site_depth),bty="n",ncol=2)
```

This graph suggests that gut-hooked fish at `Shallow` depths are the *least* likely to survive. I don't think there's enough data to support this strongly, but it is clear that the increased mortality among gut-hooked fish was greater at `Shallow` depths.
